---
title: "Compare PsychAD with SEA_AD and Mathys, et al."
subtitle: 'Compare differential expression signatures'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
    toc: true
    smart: false
vignette: >
  %\VignetteIndexEntry{SEA_AD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---





<!--- 

cd /sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/replication/
git pull 


system("git pull"); 


rmarkdown::render("compare.Rmd");



# https://hoffmg01.hpc.mssm.edu/replication/compare.html

--->



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  cache = TRUE,
  cache.lazy = FALSE)
```

# Load packages
```{r load, cache=FALSE}
library(tidyverse)
library(ggplot2)
library(ggcorrplot)
library(ggrepel)
library(parallel)
library(kableExtra)
library(arrow)
```


```{r load.data}
file = "/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/replication/Mathys_2023/Mathys_Cell_2023_Major_Cell_Type_dreamlet.parquet"
res.mathys = read_parquet(file)

file = "/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/replication/SEA_AD/SEA_AD_dreamlet.parquet"
res.seaad = read_parquet(file)

files = dir("/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/analysis/results/meta/", pattern = "topTable_meta_.*FE.parquet", full.names=TRUE)

res.pad = open_dataset(files)  %>%
   rename(logFC = "estimate", 
    stat='statistic', 
    P.Value = 'p.value') %>%
   filter(coef %in% c('m01x', 'm17x', 'm18x', 'm19x')) %>%
   collect

file = "/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/replication/celltype_mapping.tsv"
df_celltype = read_tsv( file ) 

process = function(IDs){
  sapply(IDs, function(x) strsplit(x, ",")) %>%
    unlist  %>%
    array %>%
    unique
}

cellTypeOrder = list(
PsychAD_class = df_celltype %>% filter(annotationLevel == "class") %>% pull(PsychAD),
PsychAD_subclass = df_celltype %>% filter(annotationLevel == "subclass") %>% pull(PsychAD),
SEAAD_subclass = process(df_celltype$SEA_AD_subclass),
SEAAD_supertype = process(df_celltype$SEA_AD_supertype),
Mathys_major_cell_type =process(df_celltype$Mathys_major_cell_type),
Mathys_cell_type = process(df_celltype$Mathys_cell_type))

# Given CT1 in df_celltype$PsychAD, 
# check if CT2 is a match in that row
isValidPair = function( CT1, CT2 ){

  df_sub = df_celltype %>% 
            filter(PsychAD == CT1) 

  validCellTypes = df_sub[,3:6] %>%
                  array %>%
                  unlist %>%
                  unique %>%
                  paste(collapse=",")  %>%
                  strsplit(",") %>%
                  unlist

  CT2 %in% validCellTypes
}  

```

# Microglia
```{r plotConcordance}
plotConcordance = function( tab, xlab, ylab, col = c("grey70", "#b49dd1", "#9dd0d1", "red2")){

  tab$Signif = "no"
  i = with(tab, adj.P.Val.x < 0.05)
  tab$Signif[i] = xlab
  i = with(tab, adj.P.Val.y < 0.05)
  tab$Signif[i] = ylab
  i = with(tab, adj.P.Val.x < 0.05 & adj.P.Val.y < 0.05)
  tab$Signif[i] = "both"

  tab$Signif = factor(tab$Signif, c("no", xlab, ylab, "both"))
  names(col) = levels(tab$Signif)

  # fit = lm(logFC.y ~ logFC.x, tab)
  # pv = coef(summary(fit))[2,4]
  # rvalue = format(summary(fit)$r.squared, digits=3)
  # txt = paste("R2 =", rvalue, "\np =", format(pv, digits=3))

  res = with(tab, cor.test(logFC.x, logFC.y, method="spearman"))
  rho = format(res$estimate, digits=2)
  p = format(res$p.value, digits=3)
  # txt1 = bquote(rho==.(rho))
  # txt2 = bquote(p==.(p))
  txt = paste0("rho = ", rho, "\np = ", format(p, digits=3))

  # top genes
  tab.top = tab %>% 
    arrange(-stat.x*stat.y) %>%
    filter(adj.P.Val.x < 0.05 & adj.P.Val.y < 0.05) %>%
    head(10)

  tab %>% 
    arrange(as.numeric(Signif)) %>%
    ggplot(aes(logFC.x, logFC.y, color=Signif)) +
      geom_point() +
      theme_classic() +
      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
      geom_abline(color="grey", linetype="dashed") +
      geom_hline(yintercept=0, color="grey") +
      geom_vline(xintercept=0, color="grey") +
      annotate("text", x=-.6, y=1.6, label=txt) +
      scale_color_manual(name = "FDR < 5%", values=col) +
      geom_text_repel(data = tab.top, aes(x = logFC.x, y = logFC.y, label=ID), box.padding=.6, label.padding=.6, min.segment.length=.1, color="black") +
      xlab(paste0("logFC (", xlab, ")")) +
      ylab(paste0("logFC (", ylab, ")"))
}

# Microglia: Mathys
tab = inner_join(res.pad %>% 
                  filter(AnnoLevel == "subclass", 
                    assay == 'Micro', 
                    coef == 'm01x') %>%
                  mutate(adj.P.Val = p.adjust(P.Value, "fdr")),
                  res.mathys %>% 
                  filter(assay == 'Mic', 
                    coefSet == 'ADdiag2typesAD') %>%
                  mutate(adj.P.Val = p.adjust(P.Value, "fdr")),
                  by = "ID" )

xlab = "PsychAD m01x"
ylab = "Mathys AD"

plotConcordance(tab, xlab, ylab) + ggtitle("Microglia")

# Microglia: SEA-AD
tab = inner_join(res.pad %>% 
                  filter(AnnoLevel == "subclass", 
                    assay == 'Micro', 
                    coef == 'm19x') %>%
                  mutate(adj.P.Val = p.adjust(P.Value, "fdr")),
                  res.seaad %>% 
                  filter(assay == 'Microglia-PVM', 
                    coefSet == 'CERAD.value') %>%
                  mutate(adj.P.Val = p.adjust(P.Value, "fdr")),
                  by = "ID" )

xlab = "PsychAD m19x"
ylab = "SEA-AD CERAD"

plotConcordance(tab, xlab, ylab) + ggtitle("Microglia")
```

```{r plotTiles}
plotTiles = function(df1, df2, xlab, ylab){

  grd = expand.grid(CT1 = unique(df1$assay), 
                    CT2 = unique(df2$assay) )

  df = mclapply( seq(nrow(grd)), function(i){

    tab.sub1 = df1 %>% 
            filter(assay == grd$CT1[i])

    tab.sub2 = df2 %>% 
            filter(assay == grd$CT2[i])

    tab.join = inner_join(tab.sub1, tab.sub2, by="ID")

    res = with(tab.join, cor.test(logFC.x, logFC.y, method="spearman"))

    n = nrow(tab.join)
    data.frame(CT1 = grd$CT1[i], 
              CT2 = grd$CT2[i], 
              rho = res$estimate,
              se = sqrt((1-res$estimate^2)/(n-2)),
              p.value = res$p.value)
  }, mc.cores=36)
  df = bind_rows(df)

  df$isMatch = sapply(seq(nrow(df)), function(i){
    isValidPair( df$CT1[i], df$CT2[i])
    })
 
  # make plot
  fig = df %>%
    mutate(CT1 = factor(CT1, levels(df1$assay))) %>%
    mutate(CT2 = factor(CT2, rev(levels(df2$assay)))) %>%
    ggplot(aes(CT1, CT2, fill=pmax(rho, 0))) +
      geom_tile() +
      geom_text(aes(label=ifelse(isMatch, "*", "")), size=5, hjust="middle", vjust=.9) +
      theme_classic() +
      theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust=1)) +
      coord_equal() +
      scale_fill_gradient2(name = "rho", low="blue2", mid="white", high="red2", limits=c(0, .64)) +
      xlab(xlab) +
      ylab(ylab) 

  list(fig = fig, tab = df)      
}

plotConcordanceRank = function(tab, ylab){
  main = paste("Concordance with", ylab)
  tab %>%
    mutate(CT1 = factor(CT1, rev(levels(CT1)))) %>%
    arrange(isMatch) %>%
    ggplot(aes(CT1, rho, color=isMatch)) +
      geom_point() +
      # geom_errorbar(aes(ymin = rho - 1.96*se, ymax = rho + 1.96*se), width=0) +
      theme_classic() +
      theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
      xlab(xlab) +
      coord_flip() +
      scale_color_manual(values = c("grey", "red")) +
      ylab("Spearman correlation bewteen logFC") +
      ggtitle(main)
}
```

# PsychAD class vs Mathys
## PsychAD m01x vs Mathys ADdiag2typesAD
```{r m01x_mathys_AD_class}
df1 = res.pad %>%
  filter(AnnoLevel == "class", coef == 'm01x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_class))

df2 = res.mathys %>% 
  filter(coefSet == 'ADdiag2typesAD') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m01x" 
ylab = "Mathys, et al AD"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
# res$tab %>%
#   mutate(p.value = format(p.value, digits=3, scientific=TRUE)) %>%
#     kbl(row.names=FALSE) %>%
#     kable_styling(full_width = FALSE)
```


## PsychAD m17x vs Mathys cogdx
```{r m17x_mathys_braaksc_class}
df1 = res.pad %>%
  filter(AnnoLevel == "class", coef == 'm17x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_class))

df2 = res.mathys %>% 
  filter(coefSet == 'cogdx') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m17x" 
ylab = "Mathys, et al cogdx"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```


## PsychAD m18x vs Mathys braaksc
```{r m18x_mathys_braaksc_class}
df1 = res.pad %>%
  filter(AnnoLevel == "class", coef == 'm18x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_class))

df2 = res.mathys %>% 
  filter(coefSet == 'braaksc') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m18x" 
ylab = "Mathys, et al braaksc"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```

## PsychAD m19x vs Mathys CERAD
```{r m19x_mathys_braaksc_class}
df1 = res.pad %>%
  filter(AnnoLevel == "class", coef == 'm19x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_class))

df2 = res.mathys %>% 
  filter(coefSet == 'CERAD') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m19x" 
ylab = "Mathys, et al CERAD"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```


# PsychAD subclass vs Mathys
## PsychAD m01x vs Mathys ADdiag2typesAD
```{r m01x_mathys_AD_subclass}
df1 = res.pad %>%
  filter(AnnoLevel == "subclass", coef == 'm01x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_subclass))

df2 = res.mathys %>% 
  filter(coefSet == 'ADdiag2typesAD') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m01x" 
ylab = "Mathys, et al AD"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
# res$tab %>%
#   mutate(p.value = format(p.value, digits=3, scientific=TRUE)) %>%
#     kbl(row.names=FALSE) %>%
#     kable_styling(full_width = FALSE)
```


## PsychAD m17x vs Mathys cogdx
```{r m17x_mathys_braaksc_subclass}
df1 = res.pad %>%
  filter(AnnoLevel == "subclass", coef == 'm17x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_subclass))

df2 = res.mathys %>% 
  filter(coefSet == 'cogdx') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m17x" 
ylab = "Mathys, et al cogdx"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```


## PsychAD m18x vs Mathys braaksc
```{r m18x_mathys_braaksc_subclass}
df1 = res.pad %>%
  filter(AnnoLevel == "subclass", coef == 'm18x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_subclass))

df2 = res.mathys %>% 
  filter(coefSet == 'braaksc') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m18x" 
ylab = "Mathys, et al braaksc"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```

## PsychAD m19x vs Mathys CERAD
```{r m19x_mathys_CERAD_subclass}
df1 = res.pad %>%
  filter(AnnoLevel == "subclass", coef == 'm19x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_subclass))

df2 = res.mathys %>% 
  filter(coefSet == 'CERAD') %>%
  rename(coef = coefSet) %>%
  mutate(assay = factor(assay, cellTypeOrder$Mathys_major_cell_type))

xlab = "PsychAD m19x" 
ylab = "Mathys, et al CERAD"
   
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```










# PsychAD class vs SEA-AD

## PsychAD m17x vs SEA-AD Cognitive.StatusDementia
```{r m17x_SEAAD_dementia_class}
df1 = res.pad %>%
  filter(AnnoLevel == "class", coef == 'm17x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_class))
 
df2 = res.seaad %>% 
      filter( coefSet == 'Cognitive.StatusDementia') %>%
      mutate(assay = factor(assay, cellTypeOrder$SEAAD_subclass))

xlab = "PsychAD m17x" 
ylab = "SEA-AD Cognitive.StatusDementia"
  
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```

## PsychAD m18x vs SEA-AD Braak.value
```{r m18x_SEAAD_Braak_class}
df1 = res.pad %>%
  filter(AnnoLevel == "class", coef == 'm18x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_class))
 
df2 = res.seaad %>% 
      filter( coefSet == 'Braak.value') %>%
      mutate(assay = factor(assay, cellTypeOrder$SEAAD_subclass))

xlab = "PsychAD m18x" 
ylab = "SEA-AD Braak.value"
  
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```


## PsychAD m19x vs SEA-AD CERAD
```{r m19x_SEAAD_Braak_class}
df1 = res.pad %>%
  filter(AnnoLevel == "class", coef == 'm19x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_class))
 
df2 = res.seaad %>% 
      filter( coefSet == 'CERAD.value') %>%
      mutate(assay = factor(assay, cellTypeOrder$SEAAD_subclass))

xlab = "PsychAD m19x" 
ylab = "SEA-AD CERAD.value"
  
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```




# PsychAD subclass vs SEA-AD

## PsychAD m17x vs SEA-AD Cognitive.StatusDementia
```{r m17x_SEAAD_dementia_subclass}
df1 = res.pad %>%
  filter(AnnoLevel == "subclass", coef == 'm17x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_subclass))
 
df2 = res.seaad %>% 
      filter( coefSet == 'Cognitive.StatusDementia') %>%
      mutate(assay = factor(assay, cellTypeOrder$SEAAD_subclass))

xlab = "PsychAD m17x" 
ylab = "SEA-AD Cognitive.StatusDementia"
  
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```

## PsychAD m18x vs SEA-AD Braak.value
```{r m18x_SEAAD_Braak_subclass}
df1 = res.pad %>%
  filter(AnnoLevel == "subclass", coef == 'm18x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_subclass))
 
df2 = res.seaad %>% 
      filter( coefSet == 'Braak.value') %>%
      mutate(assay = factor(assay, cellTypeOrder$SEAAD_subclass))

xlab = "PsychAD m18x" 
ylab = "SEA-AD Braak.value"
  
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```


## PsychAD m19x vs SEA-AD CERAD
```{r m19x_SEAAD_Braak_subclass}
df1 = res.pad %>%
  filter(AnnoLevel == "subclass", coef == 'm19x') %>%
  mutate(assay = factor(assay, cellTypeOrder$PsychAD_subclass))
 
df2 = res.seaad %>% 
      filter( coefSet == 'CERAD.value') %>%
      mutate(assay = factor(assay, cellTypeOrder$SEAAD_subclass))

xlab = "PsychAD m19x" 
ylab = "SEA-AD CERAD.value"
  
res = plotTiles(df1, df2, xlab, ylab)
res$fig
plotConcordanceRank( res$tab, ylab)
```












