---
title: "Analysis of NPS/AD"
subtitle: 'STAR-solo R1-4 merged h5ad'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
vignette: >
  %\VignetteIndexEntry{Decorrelate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---


<!--- 

cd /hpc/users/hoffmg01/work/nps_ad
ml python
R
# rm -rf analysis_2_cache/

system("ml git; git pull")
rmarkdown::render("analysis_2.Rmd");


# https://hoffmg01.u.hpc.mssm.edu/nps_ad/


ml git
cd ~/build2/dreamlet
git pull
R CMD INSTALL .


--->



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  cache = TRUE,
  cache.lazy = FALSE)
```

```{r load.packages, cache=FALSE}
# Use cache=FALSE so that package are fully loaded each time
# This ensures that forks within mclapply() have these loaded
# Othewise, mclapply() not have access to these libraries and will fail 
#   unless the libraries are manually loaded within each fork
suppressPackageStartupMessages({
library(SingleCellExperiment)
library(zellkonverter)
library(DelayedArray)
library(DelayedMatrixStats)
library(dreamlet)
library(muscat)
library(cowplot)
library(zenith)
library(scater)
library(HDF5Array)
library(data.table)
library(S4Vectors)
library(data.table)
})

# packageVersion("zellkonverter") >= 1.3.1
# https://github.com/theislab/zellkonverter

```


```{r load.data}
# read snRNA-seq data with annotations build in
h5ad_file = "/sc/arion/projects/CommonMind/leed62/NPS-AD/analysis_solo/NPS-AD_roundMerged_pass2_anno.h5ad"
sceCombine = readH5AD(h5ad_file, "counts", use_hdf5=TRUE)

# read metadata
meta_file = "/sc/arion/projects/CommonMind/leed62/NPS-AD/MSSM_HBCC_RUSH_clinical_metadata_combo.csv"
df_meta = fread(meta_file)

# save original order of cells
colData(sceCombine)$row.order = 1:nrow(colData(sceCombine))

# merge colData with additional metadata
df_merge = merge(colData(sceCombine), df_meta, by="SubID", all.x=TRUE)

# save merged data as colData while retaining original cell order
colData(sceCombine) = df_merge[order(df_merge$row.order),]
```

Data freeze includes 4 rounds, `r length(table(sceCombine$batch))` batches, `r length(table(sceCombine$SubID))` donors, and `r format(ncol(sceCombine), big.mark=',')` cells passing QC.

# Joint UMAP
```{r umap, dev="png"}
# extract UMAP coordinates and annotations
df = cbind(reducedDim(sceCombine, "X_umap"), colData(sceCombine)[,c("anno", "celltype", "class", "subtype", "leiden_labels")])
df = data.frame(df)

ggplot(df, aes(V1, V2, color=anno)) + geom_point(size=.05) + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2")
```

```{r umap2, dev="png"}
ggplot(df, aes(V1, V2, color=celltype)) + geom_point(size=.05) + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2")
```

```{r umap3, dev="png"}
ggplot(df, aes(V1, V2, color=class)) + geom_point(size=.05) + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2")
```

```{r umap4, dev="png"}
ggplot(df, aes(V1, V2, color=subtype)) + geom_point(size=.05) + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2")
```

```{r umap5, dev="png"}
ggplot(df, aes(V1, V2, color=leiden_labels)) + geom_point(size=.05) + theme_classic() + theme(aspect.ratio=1) + guides(colour = guide_legend(override.aes = list(size = 1.5))) + xlab("UMAP1") + ylab("UMAP2")
```


```{r combineData}
# Specify how to collapse into pseudo-bulk
sceCombine$id <- paste0(sceCombine$batch, '_', sceCombine$SubID)
sceCombine_prep <- prepSCE(sceCombine, 
    kid = "anno", # subpopulation assignments
    gid = "SubID",  # group IDs (i.e. Donor)
    sid = "id",   # sample IDs (i.e. Batch_Donor)
    drop = TRUE)

saveRDS(sceCombine_prep, file="/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/sceCombine_prep_v2.RDS" )

# Create pseudo-bulk SingleCellExperiment
pbObj <- aggregateToPseudoBulk(sceCombine_prep,
    assay = "counts", 
    fun = "sum",
    by = c("cluster_id", "sample_id"),
    BPPARAM = SnowParam(8, progressbar=TRUE))

saveRDS(pbObj, file="/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/pbObj_v2.RDS" )
```



```{r exit1, cache=FALSE, eval=FALSE}
knitr::knit_exit()
```








# Get per sample/per cellType summary of statistics
# include only included cells
```{r collapse}
df_sub = df_cell[cellID %in% colnames(sceCombine_prep),]
df_sub[,tmp:=strsplit(cellID, '_')]
df_sub[,Batch:=sapply(tmp, function(x) x[1])]
df_sub[,Donor:=sapply(tmp, function(x) x[2])]
df_sub[,barcode:=sapply(tmp, function(x) x[3])]
df_sub[,tmp:=NULL]

df_collapse = df_sub[,data.table(ID = paste0(Batch, '_', Donor),
                              n_genes = mean(n_genes),
                              n_counts = mean(n_counts),
                              percent_mito = mean(percent_mito),
                              scale = mean(scale),
                              UMAP_1 = mean(UMAP_1),
                              UMAP_2 = mean(UMAP_2)),
                              by=c('Batch', 'Donor', 'anno')]
```


```{r voom}
# Normalize and apply voom
res.proc = processAssays( pbObj, ~ 1,#(1|Donor) + (1|Batch), 
  min.cells = 5,
  min.count = 10, 
  pmetadata = df_collapse,  
  pkeys=c('ID', 'anno'),
  BPPARAM = SnowParam(16, progressbar=TRUE))
  
res.proc
```



```{r voom.plot, fig.height=10, fig.width=10, cache=FALSE}
plotVoom( res.proc )
```


```{r voom.save, fig.height=10, fig.width=10, cache=FALSE}
saveRDS( res.proc, "/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/res.proc.RDS")
```


```{r show.chrXY, eval=FALSE}

df_sex = lapply( names(res.proc), function(CT){
  geneExpr = res.proc[[CT]]
  df = NULL
  if( sum(c("XIST", "UTY") %in% rownames(geneExpr)) == 2){  
    df = data.frame(cellType = CT, IDS = colnames(geneExpr$E), XIST = geneExpr$E["XIST",], UTY = geneExpr$E["UTY",])
  }
  df
})
df_sex = do.call(rbind, df_sex)
df_sex = data.table(df_sex)

fig = ggplot(df_sex, aes(XIST, UTY)) + geom_point() + theme_classic() + facet_wrap(~cellType)
ggsave(file="test.pdf", fig)

df2 = df_sex[,data.table(Xmax = max(XIST), Ymax = max(UTY)), by="IDS"]
df2[,isMale:=(Ymax > 7.5) & (Xmax < 10.5)]
df2[,Sex:=ifelse(isMale, "M", "F")]

saveRDS(df2, file="df_sex.RDS")


fig = ggplot(df2, aes(Xmax, Ymax, color=Sex)) + geom_point() + theme_classic() + geom_abline(slope=.8)
ggsave(file="test.pdf", fig, height=6, width=6)


colData(res.proc) = merge(colData(res.proc), df2[,c("IDS", "Sex")], by.x="row.names", by.y="IDS", all.x=TRUE)


df_link = fread("/sc/arion/projects/psychAD/NPS-AD/freeze1_rc/metadata/NPS-AD_roundMerged_meta.csv")

df_link2 = fread("/sc/arion/projects/psychAD/upload2synpase/final_NPSAD_snRNAseq_metrics_combo.csv")


df_link[1:3,]
df_link2[1:3, 1:10]
colData(sceCombine_filter)$id2 = sapply(strsplit( colData(sceCombine_filter)$id,"_"), function(x) x[1])


head(colData(sceCombine_filter))$id[1:4]

df_merge = merge(colData(sceCombine_filter)[1:10000,], df_link2, by.x="id2", by.y="HTO_ID", all.x=TRUE)
dim(df_merge)


h5ad_file = "/sc/arion/projects/CommonMind/leed62/NPS-AD/analysis_solo/NPS-AD_roundMerged_pass2_anno.h5ad"

sce = readH5AD(h5ad_file, "counts", use_hdf5=TRUE)


df_meta = fread("/sc/arion/projects/CommonMind/leed62/NPS-AD/MSSM_HBCC_RUSH_clinical_metadata_combo.csv")

df_merge = merge(colData(sce), df_meta, by="SubID", all.x=TRUE, sort=FALSE)
dim(df_merge)






```





```{r dreamlet.fit}
fit.dl = dreamlet( res.proc, ~ (1|Batch), BPPARAM=SnowParam(36))

saveRDS( fit.dl, "/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/fit.dl.RDS")
```




```{r varPart}
res.vp = fitVarPart(res.proc, ~ (1|Donor) + (1|Batch), BPPARAM = SnowParam(18, progressbar=TRUE))
```

```{r vp.plot, fig.height=10, fig.width=10, cache=FALSE}
plotVarPart( sortCols(res.vp) )
```




### Examine variancePartition results
```{r vp.examine, fig.height=2, fig.width=7}
CT = "Astrocyte"

i = 1:3
df = sortCols(res.vp)[res.vp$assay == CT,]
rownames(df) = df$gene
i = c(i, which.max(df$Donor), which.max(df$Batch), which.max(df$Residuals))

plotPercentBars( data.frame(df[i,-c(1,2)])) + ggtitle(CT)
```

Examples

```{r vp.examine2, fig.height=6, fig.width=8, cache=TRUE}      
figList = lapply( df$gene[i], function(geneName){
  df = merge( as.data.frame(t(assay(res.proc, CT)$E[geneName,,drop=FALSE])), colData(res.proc), by="row.names")

  plotStratifyBy(df, 'group_id', geneName, colorBy=NULL, main=geneName) + theme(aspect.ratio=1)
})
plot_grid( plotlist=figList, nrow=2 )
```


# Correlation with and between donors
```{r within_btw, fig.height=10, fig.width=10}
source("/sc/arion/work/hoffmg01/nps_ad/common_analyses.R")

eval_within_across_donor( res.proc )
```


```{r varPart2}
res.vp2 = fitVarPart(res.proc, ~ (1|Donor) + (1|Batch) + n_genes + n_counts + percent_mito    scale + UMAP_1 + UMAP_2, BPPARAM = SnowParam(18, progressbar=TRUE))
```

```{r vp.plot2, fig.height=10, fig.width=10, cache=FALSE}
plotVarPart( sortCols(res.vp2) )
```





# Outliers
```{r outrider}

lapply(fit.dl, function(fit){
  resid = residuals(fit)




  })




```

<!---
plotCellComposition()

cellTypeCompositionTest()
--->

TODO

 - add 'mega-batches' 1,2,3
 - add UMAP coordinates
 - cell composition fractions
   - variancePartition in composition
 - how does cell number, read count influence within donor correlation
   - What happens when I remove batch effect?



show correlation within/between batch
Get collapsed experssion per donor



```{r exit, cache=FALSE, eval=FALSE}
knitr::knit_exit()
```











