---
title: "Analysis of NPS/AD"
subtitle: 'Loss of chrY'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
vignette: >
  %\VignetteIndexEntry{Decorrelate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---


<!--- 

cd /sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/
ml python git
git pull
R --vanilla

system("git pull")
rmarkdown::render('chrom_loss.Rmd');




# https://hoffmg01.u.hpc.mssm.edu/nps_ad/


# submit as job

DIR=`pwd`
echo '#!/bin/bash' > loss_of_Y.lsf 
echo "#BSUB -J loss_of_Y
#BSUB -P acc_CommonMind
#BSUB -q premium
#BSUB -n 36
#BSUB -R span[hosts=1]
#BSUB -W 12:00
#BSUB -o $DIR/loss_of_Y_%J.stdout
#BSUB -eo $DIR/loss_of_Y_%J.stderr
#BSUB -L /bin/bash

ml R/4.1.0-cairo libpng/16 python

R_LIBS_USER=/hpc/users/hoffmg01/.Rlib/R_410/
R_LIBS=$R_LIBS_USER:$R_LIBS

cd /hpc/users/hoffmg01/work/nps_ad

echo \"rmarkdown::render('loss_of_Y.Rmd')\" | R --vanilla
" >> loss_of_Y.lsf 

bsub < loss_of_Y.lsf

ml git
cd ~/build2/dreamlet
git pull
R CMD INSTALL .


--->

Examine distribution of chromosome loss events per cell 
  Is the probability of losing chrA related to loss of chrB after correcting for depth
  show overall loss rate per chromosome vs depth
  Is count per cell Poisson distributed?
  Use t-statistic as a proxy?


Donor level
  ashr within each cell type
  loss with age
  relate to phenotype

Collapse GABA neurons
  out-of-sample prediction



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

```{r load.packages, cache=FALSE}
# Use cache=FALSE so that package are fully loaded each time
# This ensures that forks within mclapply() have these loaded
# Othewise, mclapply() not have access to these libraries and will fail 
#   unless the libraries are manually loaded within each fork
library(SingleCellExperiment)
library(zellkonverter)
library(DelayedArray)
library(DelayedMatrixStats)
library(HDF5Array)
library(dreamlet)
library(scater)
library(OUTRIDER)
library(tidyverse)
library(cowplot)
library(kableExtra)
library(GenomicRanges)
source("./helper/qqplot.R")
```


```{r load.data}
datafile = "/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/sceCombine_f1.RDS"

sceCombine = readRDS( datafile )

# collapse GABAergic neurons
sceCombine$celltype8 = sceCombine$celltype %>%
    as.character %>% 
    recode( 
    "GABAergic SST interneuron" = 'GABAergic neuron',
    "GABAergic VIP interneuron" = 'GABAergic neuron',
    "GABAergic PVALB interneuron" = 'GABAergic neuron') %>% 
    as.factor

chromExpr = rowsum(assay(sceCombine, "X"), rowData(sceCombine)$gene_chrom)
```

# OUTRIDER analysis
```{r outrider.v2}
# Include everything
ods.all = OutriderDataSet(countData=chromExpr, colData = colData(sceCombine))

ods <- filterExpression(ods.all, minCounts=TRUE, filterGenes=TRUE)

# Without autoencoder
ods = estimateSizeFactors(ods)
ods = controlForConfounders(ods, q=2, implementation="pca", BPPARAM=SnowParam(12, progressbar=TRUE))
ods = fit(ods, BPPARAM=SnowParam(1, progressbar=TRUE))
ods = computePvalues(ods, method="BH")
ods = computeZscores(ods)

res = results(ods, all=TRUE)
```




```{r outrider.v2.males}
# just males
idx = which(ods.all$Sex == "Male")

ods_male <- filterExpression(ods.all[,idx], minCounts=TRUE, filterGenes=TRUE)

# Without autoencoder
ods_male = estimateSizeFactors(ods_male)
ods_male = controlForConfounders(ods_male, q=2, implementation="pca", BPPARAM=SnowParam(12, progressbar=TRUE))
ods_male = fit(ods_male, BPPARAM=SnowParam(1, progressbar=TRUE))
ods_male = computePvalues(ods_male, method="BH")
ods_male = computeZscores(ods_male)

res_male = results(ods_male, all=TRUE)
```


```{r outrider.v2.females}
# just males
idx = which(ods.all$Sex == "Female")

ods_female <- filterExpression(ods.all[,idx], minCounts=TRUE, filterGenes=TRUE)

# Without autoencoder
ods_female = estimateSizeFactors(ods_female)
ods_female = controlForConfounders(ods_female, q=2, implementation="pca", BPPARAM=SnowParam(12, progressbar=TRUE))
ods_female = fit(ods_female, BPPARAM=SnowParam(12, progressbar=TRUE))
ods_female = computePvalues(ods_female, method="BH")
ods_female = computeZscores(ods_female)

res_female = results(ods_female, all=TRUE)
```

```{r combine}
# combine 3 analyses
res_combine = rbind(res[res$geneID %in% c(1:22, "MT"),],
                    res_male[res_male$geneID %in% c("X", "Y"),],
                    res_female[res_female$geneID %in% c("X"),])

res_combine$padjust = p.adjust(res_combine$pValue, "fdr")
res_combine$aberrant[] = FALSE
res_combine$aberrant[res_combine$padjust < 0.05] = TRUE

outfolder = '/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/chrom_loss_results/'
file = paste0(outfolder, "outliers.RDS")
saveRDS(res_combine, file=file)
```


# Plots from main analysis of all samples
```{r plotExpressedGenes, eval=FALSE, dev="png"}
plotExpressedGenes(ods)
```


```{r plotAberrantPerSample, eval=FALSE, dev="png"}
plotAberrantPerSample(ods, padjCutoff=0.05) + theme_classic() + theme(aspect.ratio=1)
```

```{r plotQQ, fig.width=10, fig.height=35, dev="png", eval=FALSE}
figList = lapply( c(1:22, "MT", "X", "Y"), function(x){
   
  i = which(res_combine$geneID == x)
  qqplot(res_combine$pValue[i], pmin=1e-100) + ggtitle(x)
 })
plot_grid(plotlist=figList, ncol=3)
```


```{r plotVolcano, fig.width=8, fig.height=8, eval=FALSE}
for( x in rownames(ods) ){
  plotVolcano(ods, geneID=x, basePlot=TRUE) + theme_classic() + theme(aspect.ratio=1)
}
```

```{r plotExpressionRank, fig.width=8, fig.height=8, dev="png"}
for( x in rownames(ods) ){
  plotExpressionRank(ods, x, basePlot=TRUE) + theme_classic() + theme(aspect.ratio=1)
}
```

```{r plotExpectedVsObservedCounts, fig.height=12, fig.width=8, dev="png"}
source("./helper/plotExpObsCounts.R")
plotExpObsCounts(ods, "21")
plotExpObsCounts(ods, "Y")
# figList = lapply(rownames(ods), function(x){
#   plotExpObsCounts(ods, x)  
# })
# plot_grid(plotlist=figList, ncol=3)
# rm(figList); gc()
```


```{r plotPowerAnalysis, fig.width=5, fig.height=4}
plotPowerAnalysis(ods) + theme_classic() + theme(aspect.ratio=1)
```


# Analysis per cell type
```{r per celltype}

run_by_celltype = function(chromExprObj, sce){

  res_CT = lapply( levels(colData(sce)$celltype8), function(CT){

    message(CT)
    idx = colData(sce)$celltype8 == CT

    ods = OutriderDataSet(countData=chromExprObj[,idx], 
                          colData = colData(sce)[idx,])

    ods <- filterExpression(ods, minCounts=TRUE, filterGenes=TRUE)

    # Without autoencoder
    ods = estimateSizeFactors(ods)
    ods = controlForConfounders(ods, q=2, implementation="pca", BPPARAM=SerialParam())
    ods = fit(ods, BPPARAM=SerialParam())
    ods = computePvalues(ods, method="BH", BPPARAM=SerialParam())
    ods = computeZscores(ods, BPPARAM=SerialParam())

    res = results(ods, all=TRUE)
    res$celltype = CT
    res
  })
  do.call(rbind, res_CT)
}

resAll = run_by_celltype(chromExpr, sceCombine)

idx = colData(sceCombine)$Sex == "Male"
resMale = run_by_celltype(chromExpr[,idx], sceCombine[,idx])

idx = colData(sceCombine)$Sex == "Female"
resFemale = run_by_celltype(chromExpr[,idx], sceCombine[,idx])

# combine 3 analyses
res_combine = rbind(resAll[res$geneID %in% c(1:22, "MT"),],
                    resMale[resMale$geneID %in% c("X", "Y"),],
                    resFemale[resFemale$geneID %in% c("X"),])

res_combine$padjust = p.adjust(res_combine$pValue, "fdr")
res_combine$aberrant[] = FALSE
res_combine$aberrant[res_combine$padjust < 0.05] = TRUE

outfolder = '/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/chrom_loss_results/'
file = paste0(outfolder, "outliers_by_celltype8.RDS")
saveRDS(res_combine, file=file)
```









