---
title: "Resample rho"
subtitle: 'Automated analysis'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
output: 
  html_document:
    toc: true
    smart: true
    self_contained: false
params:
  dataset: NULL
  variable_type: NULL
  ctst_key: NULL
  AnnoLevel: NULL
  SampleLevel: NULL
---

<!---


cd /sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/analysis/freeze2/revision
git pull

system("git pull")

rmarkdown::render("sample_rho.Rmd")

--->



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = FALSE,
  cache.lazy = FALSE)
```

```{r plots}
library(tidyverse)
library(ggrepel)
file = "/sc/arion/projects/psychAD/NPS-AD/freeze2_proc/231229_PsychAD_capstone_F4_5_crossDis/241202_PsychAD_SharedHeritability-TranscriptomeSimilarity_export.csv"
df = read_csv(file) %>%
  rename(hsq = Herit_estimate, hsq.se = Herit_SE,
      geneExpr = Simil_mean, geneExpr.se = Simil_SE)
colnames(df)[1] = "label" 

col = df$pal
names(col) = df$type

rho = with(df, cor(hsq, geneExpr, method="spearman"))
rho = format(rho, digits=3)

ggplot(df, aes(hsq, geneExpr)) +
  geom_errorbar(aes(ymin = geneExpr - 1.96*geneExpr.se, ymax = geneExpr + 1.96*geneExpr.se, color=type), alpha = .3) +
  geom_errorbar(aes(xmin = hsq - 1.96*hsq.se, xmax = hsq + 1.96*hsq.se, color=type), alpha = .3) +
  geom_point(aes(color=type)) +
  geom_smooth(method="lm") +
  xlab("Shared heritability") +
  ylab("Transcriptome similarity") +
  theme_classic() +
  theme(aspect.ratio=1, plot.title = element_text(hjust = 0.5)) +
  geom_text_repel(aes(label=label)) +
  scale_color_manual(values=col) +
  coord_fixed(xlim=c(-.2, .75)) +
  ggtitle(bquote(rho == .(rho)))
```

Errors bars indicate 95% confidence interval

```{r resample, fig.width=5, fig.height=5}
rhohat = sapply(seq(10000), function(i){
  x1 = with(df, rnorm(length(hsq), hsq, hsq.se))
  x2 = with(df, rnorm(length(geneExpr), geneExpr, geneExpr.se))
  cor(x1, x2, method="spearman")
})

pnorm(0, mean(rhohat), sd(rhohat))

par(pty="s")
hist(rhohat, xlab="Estimate of rho", main="Distribution of rho drawn from synthetic data")
abline(v=0, col="red", lwd=2, lty=2)
```

The association betweeen estimated transcriptome similarity and shared heritability values was estimated using an approach to model the uncertainty about these values.  Synthetic values were simulated from normal distribution centered at each estimate with the corresponding standard error.  For each of 10,000 synthetic datasets, the Spearman correlation between the sampled transcriptome similarity and shared heritability was computed.  Using a normal approximation of the rho statistic, we can reject the null hypothesis that rho is less than zero with p-value `r pnorm(0, mean(rhohat), sd(rhohat))`.


