---
title: "Evaluate concordance based on formula"
subtitle: 'Automated analysis'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
output: 
  html_document:
    toc: true
    smart: true
    self_contained: false
params:
  dataset: NULL
  variable_type: NULL
  ctst_key: NULL
  AnnoLevel: NULL
  SampleLevel: NULL
---



<!--- 

# https://hoffmg01.hpc.mssm.edu/nps_ad/analysis/freeze2/testing/compare_concordace.html

cd /sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/analysis/freeze2/testing/
ml python git pandoc gcc/11.2.0
git pull origin master

rm -rf compare_concordace_cache/ compare_concordace_files

system("git pull origin master")
rmarkdown::render("compare_concordace.Rmd")


--->


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

```{r libs, cache=FALSE}
library(dreamlet)
library(tidyverse)
library(parallel)
library(RhpcBLASctl)
omp_set_num_threads(2)
```


# CMC Signatures
```{r cmc}
file = "/sc/arion/projects/CommonMind/hoffman/cmc_signature/df_DE_results.RDS"
res.cmc = readRDS(file)

df.cmc = lapply(names(res.cmc), function(x){
  res.cmc[[x]]$DxSCZ %>%
    as_tibble %>%
    mutate(estimate = logFC, 
        std.error = logFC / t, 
        ID = Symbol, 
        statistic = t,
        p.value = P.Value,
        FDR.within =  adj.P.Val,
        assay = x) %>%
    select(ID, assay, estimate, std.error, statistic, p.value, FDR.within)
  })
df.cmc = bind_rows(df.cmc)

df.cmc$assay[df.cmc$assay == 'MSSM-Penn-Pitt'] = "MPP"
df.cmc$assay[df.cmc$assay == 'NIMH-HBCC'] = "HBCC"

tab = table(df.cmc$ID)
exclude = names(tab[tab !=2])

df_filter = df.cmc %>%
  select(ID, estimate, assay) %>%
  filter(! ID %in% exclude) %>%
  pivot_wider(names_from = assay, values_from = estimate) 

df_filter_p = df.cmc %>%
  select(ID, p.value, assay) %>%
  filter(! ID %in% exclude) %>%
  pivot_wider(names_from = assay, values_from = p.value)

df_filter_se = df.cmc %>%
  select(ID, std.error, assay) %>%
  filter(! ID %in% exclude) %>%
  pivot_wider(names_from = assay, values_from = std.error)

cor(df_filter[,-1] %>% as.matrix)
```

# PsychENCODE
```{r pec}
file = "/sc/arion/projects/CommonMind/mpjanic/PEC/final_dl/final/scz.csv"
df_pec = read.table(file, sep=',', row.names=1) %>%
      as_tibble 

df = df_pec %>%
        select(assay, estimate, ID) %>%
        mutate(assay = paste("pec/scz", assay)) %>%
        pivot_wider(names_from = "assay", values_from="estimate")

df_p = df_pec %>%
        select(assay, p.value, ID) %>%
        mutate(assay = paste("pec/scz", assay)) %>%
        pivot_wider(names_from = "assay", values_from="p.value")

df_se = df_pec %>%
        select(assay, std.error, ID) %>%
        mutate(assay = paste("pec/scz", assay)) %>%
        pivot_wider(names_from = "assay", values_from="std.error")

df_filter = df_filter %>%
            inner_join(df, by='ID')

df_filter_p = df_filter_p %>%
            inner_join(df_p, by='ID')

df_filter_se = df_filter_se %>%
            inner_join(df_se, by='ID')
```

# MORE PAIRS
```{r pairs}
library(arrow)
library(tidyverse)
library(parallel)

get_pi1 = function(p){
  library(qvalue)
  res <- tryCatch( pi0est(c(p, 1)), error = function(e) NULL)

  ifelse(is.null(res), NA, 1 - res$pi0)
}

file = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/analysis/results/topTable_combined.parquet"
df_all = read_parquet(file) %>%
          filter(AnnoLevel == "class" | AnnoLevel == "bulk") %>%
          rename(estimate = logFC)

file = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/analysis/results/meta/res_meta.parquet"
df_meta = read_parquet(file) %>%
          filter(AnnoLevel == "class" | AnnoLevel == "bulk", 
            method == "FE") %>%
          rename(P.Value = p.value)

df_stack = bind_rows(df_all, df_meta)

df_coef = df_all %>%
  select(coef, Dataset) %>%
  distinct

df_coef[grep("SC", df_coef$coef),]

getCohort = function(x){
  res = df_coef[grep(x, df_coef$coef),]$Dataset[1]
  ifelse(is.na(res), "meta-analysis", res)
}


make_plot = function(coef1, coef2){
  df1 = df_stack %>%
          filter(coef == coef1)%>%
          arrange(assay)

  df2 = df_stack %>%
          filter(coef == coef2) %>%
          arrange(assay)

  df_join = inner_join(df1, df2, by=c("assay", "ID"))

  # corr along diagonals
  # df_join %>%
  #   group_by(assay) %>%
  #   summarize(r = cor(estimate.x, estimate.y, method="sp"), 
  #             n = length(estimate.x),
  #             se = sqrt((1-r^2)/ (n-2)))

  Y1 = df1 %>%
        select(assay, estimate, ID) %>%
        pivot_wider(names_from = "assay", values_from="estimate") 

  Y2 = df2 %>%
        select(assay, estimate, ID) %>%
        pivot_wider(names_from = "assay", values_from="estimate")

  Y = inner_join(Y1, Y2, by="ID") %>%
        left_join(df_filter, by="ID") %>%
        select(-ID) 

  C = cor(Y, use="pair", method="sp" )

  main = paste0("x: ", coef1, " (", getCohort(coef1), ") | y: ", coef2, " (", getCohort(coef2), ")")

  # png("~/www/test.png", height=600, width=600)
  breaksList = seq(-1, 1, by=.025)

  pheatmap::pheatmap(C, 
    cluster_rows=FALSE, 
    cluster_cols=FALSE, 
    color=colorRampPalette(c("navy", "white", "red"))(length(breaksList)),
    breaks=breaksList, 
    main=main)
  # dev.off()
}
```


```{r plot.pairs, fig.height=7, fig.width=7, cache=FALSE}
# meta-analysis of AD/control and SCZ control
coef1 = "m01x"
coef2 = "m05x" 
make_plot(coef1, coef2)

coef1 = "c47xSCZ_crossDis - c47xControls_crossDis" # MSSM
coef2 = "c48xSCZ_crossDis - c48xControls_crossDis" # RUSH
make_plot(coef1, coef2)

coef1 = "m10x" # AD
coef2 = "m11x" # SCZ
make_plot(coef1, coef2)

coef1 = "c07xSCZ - c07xControl" # MSSM
coef2 = "c03xAD - c03xControl" # RUSH
make_plot(coef1, coef2)

coef1 = "c02xAD - c02xControl" # MSSM
coef2 = "c03xAD - c03xControl" # RUSH
make_plot(coef1, coef2)

coef1 = "c02xAD - c02xControl" # MSSM
coef2 = "c08xSCZ - c08xControl" # HBCC
make_plot(coef1, coef2)

coef1 = "c38xSCZ - c38xBD" # HBCC
coef2 = "c06xAD - c06xSCZ" # MSSM
make_plot(coef1, coef2)

coef1 = "c07xSCZ - c07xControl" # MSSM
coef2 = "c02xAD - c02xControl"  # MSSM
make_plot(coef1, coef2)

coef1 = "c03xAD - c03xControl" # RUSH
coef2 = "c08xSCZ - c08xControl" # HBCC
make_plot(coef1, coef2)


coef1 = "r10x.L" # AGING
coef2 = "c02xAD - c02xControl"  # MSSM
make_plot(coef1, coef2)

coef1 = "r10x.L" # AGING
coef2 = "m01x"  # MSSM
make_plot(coef1, coef2)

coef1 = "c08xSCZ - c08xControl" # HBCC
coef2 = "c07xSCZ - c07xControl" # MSSM
make_plot(coef1, coef2)

coef1 = "m05x" # SCZ
coef2 = "m20x" # APOE
make_plot(coef1, coef2)

coef1 = "m01x" # AD
coef2 = "m20x" # APOE
make_plot(coef1, coef2)
```

```{r plot.pi1}
make_plot = function(coef1, coef2){
  df1 = df_stack %>%
          filter(coef == coef1)%>%
          arrange(assay)

  df2 = df_stack %>%
          filter(coef == coef2) %>%
          arrange(assay)

  Y1 = df1 %>%
        select(assay, P.Value, ID) %>%
        pivot_wider(names_from = "assay", values_from="P.Value") 

  Y2 = df2 %>%
        select(assay, P.Value, ID) %>%
        pivot_wider(names_from = "assay", values_from="P.Value")

  Y = inner_join(Y1, Y2, by="ID") %>%
        left_join(df_filter_p, by="ID") %>%
        select(-ID) 

  df_grid = expand.grid(assay1 = colnames(Y), assay2 = colnames(Y)) 

  df_pi1 = mclapply( 1:nrow(df_grid), function(i){

    k1 = df_grid$assay1[i] %>% as.character
    k2 = df_grid$assay2[i]%>% as.character

    df = Y[,c(k1, k2)]
    df = df[apply(df,1, function(x) all(!is.na(x))),]

    if( k1 == k2){      
      pi1 = get_pi1(df[,1] %>% pull)
      j = p.adjust(df[,1] %>% pull, "fdr") < 0.05
    }else{
      j = p.adjust(df[,1] %>% pull, "fdr") < 0.05

      pvals = df[j,2] %>% pull

      pi1 = get_pi1(pvals)
    }

    data.frame(assay1 = k1, assay2 = k2, pi1 =pi1, nDE = sum(j) )
    }, mc.cores=12)
  df_pi1 = bind_rows(df_pi1)

  main = paste0("x: ", coef1, " (", getCohort(coef1), ") | y: ", coef2, " (", getCohort(coef2), ")")

  df_pi1 %>%
      mutate(assay1 = factor(assay1, colnames(Y))) %>%
      mutate(assay2 = factor(assay2, colnames(Y))) %>%
       ggplot(aes(assay1, assay2, fill=pi1, label=round(pi1,2))) +
            geom_tile() +
            geom_text(size=1) +
            theme_classic() +
            theme(aspect.ratio=1, 
              legend.position = "right", 
              axis.text.x = element_text(angle = 90, hjust=1,vjust=.5),
              plot.title = element_text(hjust = 0.5)) +
            scale_fill_gradient(low="white", high="red", limits=c(0,1)) +
            ggtitle(main) + 
            xlab("Discovery") +
            ylab("Replication") 
}
```



```{r plot.pairs.pi1, fig.height=7, fig.width=7, cache=FALSE}
# meta-analysis of AD/control and SCZ control
coef1 = "m01x"
coef2 = "m05x" 
make_plot(coef1, coef2)

coef1 = "c47xSCZ_crossDis - c47xControls_crossDis" # MSSM
coef2 = "c48xSCZ_crossDis - c48xControls_crossDis" # RUSH
make_plot(coef1, coef2)

coef1 = "m10x" # AD
coef2 = "m11x" # SCZ
make_plot(coef1, coef2)

coef1 = "c07xSCZ - c07xControl" # MSSM
coef2 = "c03xAD - c03xControl" # RUSH
make_plot(coef1, coef2)

coef1 = "c02xAD - c02xControl" # MSSM
coef2 = "c03xAD - c03xControl" # RUSH
make_plot(coef1, coef2)

coef1 = "c02xAD - c02xControl" # MSSM
coef2 = "c08xSCZ - c08xControl" # HBCC
make_plot(coef1, coef2)

coef1 = "c38xSCZ - c38xBD" # HBCC
coef2 = "c06xAD - c06xSCZ" # MSSM
make_plot(coef1, coef2)

coef1 = "c07xSCZ - c07xControl" # MSSM
coef2 = "c02xAD - c02xControl"  # MSSM
make_plot(coef1, coef2)

coef1 = "c03xAD - c03xControl" # RUSH
coef2 = "c08xSCZ - c08xControl" # HBCC
make_plot(coef1, coef2)


coef1 = "r10x.L" # AGING
coef2 = "c02xAD - c02xControl"  # MSSM
make_plot(coef1, coef2)

coef1 = "r10x.L" # AGING
coef2 = "m01x"  # MSSM
make_plot(coef1, coef2)

coef1 = "c08xSCZ - c08xControl" # HBCC
coef2 = "c07xSCZ - c07xControl" # MSSM
make_plot(coef1, coef2)

coef1 = "m05x" # SCZ
coef2 = "m20x" # APOE
make_plot(coef1, coef2)

coef1 = "m01x" # AD
coef2 = "m20x" # APOE
make_plot(coef1, coef2)
```
```{r exit2, cache=FALSE }
knitr::knit_exit()
```




coef1 = "c08xSCZ - c08xControl" # HBCC
coef2 = "c07xSCZ - c07xControl" # MSSM

# beta
df1 = df_stack %>%
          filter(coef == coef1)%>%
          arrange(assay)

df2 = df_stack %>%
        filter(coef == coef2) %>%
        arrange(assay)

Y1 = df1 %>%
      select(assay, estimate, ID) %>%
      pivot_wider(names_from = "assay", values_from="estimate") 

Y2 = df2 %>%
      select(assay, estimate, ID) %>%
      pivot_wider(names_from = "assay", values_from="estimate")

Beta = inner_join(Y1, Y2, by="ID") %>%
      left_join(df_filter, by="ID") 


df1 = df_stack %>%
          filter(coef == coef1)%>%
          arrange(assay)

df2 = df_stack %>%
        filter(coef == coef2) %>%
        arrange(assay)

Y1 = df1 %>%
      mutate(se = estimate / t)  %>%
      select(assay, se, ID) %>%
      pivot_wider(names_from = "assay", values_from="se") 

Y2 = df2 %>%
      mutate(se = estimate / t)  %>%
      select(assay, se, ID) %>%
      pivot_wider(names_from = "assay", values_from="se")

SE = inner_join(Y1, Y2, by="ID") %>%
      left_join(df_filter, by="ID")


identical(rownames(Beta), rownames(SE))
identical(colnames(Beta), colnames(SE))

library(heivr)

hres = heivrPairs(as.matrix(Beta[,-1]), 
                  as.matrix(SE[,-1])^2)


main = paste0("x: ", coef1, " (", getCohort(coef1), ") | y: ", coef2, " (", getCohort(coef2), ")")

png("~/www/test.png", height=600, width=600)
breaksList = seq(-1, 1, by=.025)

pheatmap::pheatmap(hres, 
  cluster_rows=FALSE, 
  cluster_cols=FALSE, 
  color=colorRampPalette(c("navy", "white", "red"))(length(breaksList)),
  breaks=breaksList, 
  main=main)
dev.off()






# Meta data
```{r metadata}
library(dreamlet)
library(tidyverse)

file = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/processAssays/final/FULL_2024-02-01_18_49_processAssays_SubID_class.RDS"
# res.proc = readRDS(file)

file = "/sc/arion/projects/CommonMind/hoffman/cmc_signature/vobj.lst.RDS"
vobj.lst = readRDS(file)

info1 = colData(res.proc) %>%
          tibble(SubID = rownames(colData(res.proc)), .) %>%
          mutate(Institution = fct_recode(Source, "MSSM" = "M", "HBCC" = "H")) %>%
          filter( Institution %in% c("MSSM", "HBCC")) %>%
          mutate(Institution = droplevels(Institution)) %>%
           mutate(Dx = c(apply(cbind(as.character(c07x), as.character(c08x)), 1, function(x){
            res = x[!is.na(x)]
            ifelse(length(res)==0, "NA", res)
            }))) %>%
          select(SubID, Institution, Age, Sex, PMI, pH, Dx)

with(info1, table(Dx, Institution))     

boxplot(Age ~ Institution*Dx, info1)     

file = "/sc/arion/projects/CommonMind/hoffman/cmc_signature/CMC_Human_clinical_metadata.csv"
info2 = read_csv(file) %>%          
          mutate(Institution = fct_recode(Institution, "HBCC" = "NIMH-HBCC")) %>%
          filter(Dx %in% c("SCZ", "Control"), 
            Institution %in% c("MSSM", "HBCC")) %>%
          mutate(Institution = droplevels(Institution)) %>%
          mutate(Age = as.numeric(`Age of Death`),
          PMI = `PMI (in hours)`,
          Sex = fct_recode(Sex, 'Male' = 'XY', "Female" = "XX"),
          SubID = `Individual ID` ) %>%
          select(SubID, Institution, Age, Sex, PMI, pH, Dx)


df = inner_join(info1, info2, by = c("Sex", 'Age', "PMI", "Institution")) %>% 
  filter(Age > 17)


tab1 = table(df$SubID.x)
u1 = names(tab1)[tab1==1]

tab2 = table(df$SubID.y)
u2 = names(tab2)[tab2==1]

df2 = df %>% 
  filter(SubID.x %in% u1, SubID.y %in% u2) %>%
  mutate(Dx.x = as.character(Dx.x),
        Dx.y = as.character(Dx.y))

df2$Dx.x[is.na(df2$Dx.x)] = "NA"
df2$Dx.y[is.na(df2$Dx.y)] = "NA"

# Dx.x = PsychAD
# Dx.y = CommonMind 
with(df2, table(Dx.x, Dx.y))


with(df2, xtabs(~ Dx.x + Dx.y + Institution))

Y_MSSM = vobj.lst[[1]]$E[,df2$SubID.y]
Y_HBCC = vobj.lst[[2]]$E[,df2$SubID.y]



```
















# V1
```{r main}
# read PB
files = c(MSSM = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/pseudobulk/MSSM_2024-02-01_16_17_PB_SubID_class.RDS", 
  HBCC = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/pseudobulk/HBCC_2024-02-01_15_18_PB_SubID_class.RDS")
pbList = lapply( files, readRDS)
names(pbList) = names(files)

# read PB bulk
# files = c(MSSM = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/pseudobulk/MSSM_2024-02-01_16_17_PB_SubID_bulk.RDS", 
#   HBCC = "/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/pseudobulk/HBCC_2024-02-01_15_18_PB_SubID_bulk.RDS")
# pbList2 = lapply( files, readRDS)
# names(pbList2) = names(files)

# assay(pbList[["MSSM"]], "bulk") = assay(pbList2[["MSSM"]], "bulk")
# assay(pbList[["HBCC"]], "bulk") = assay(pbList2[["HBCC"]], "bulk")

# for( dataset in names(pbList)){
#   for( id in names(int_colData(pbList[[dataset]])$n_cells)){
#     int_colData(pbList[[dataset]])$n_cells[[id]] = append(int_colData(pbList[[dataset]])$n_cells[[id]], c(bulk=100))
#   }
# }

# permute
# for( id in names(pbList) ){
#   v = ifelse(id == "MSSM", "c02x", "c08x")
#   values = colData(pbList[[id]])[[v]]
#   colData(pbList[[id]])[[v]] = sample(values)
# }

# processAssays
form = ~ Sex + scale(Age) + scale(PMI) + pH + log(n_genes) + percent_mito + mito_genes + ribo_genes + mito_ribo 

res.procList = mclapply(pbList, function(pb){
  # res = processAssays(pb, ~1)

  # keep = outlierByAssay(res, nPC=5) %>%
  #   group_by(ID) %>%
  #   summarize(z2 = sort(z, decreasing=TRUE)[2]) %>%
  #   arrange(desc(z2)) %>%
  #   filter(z2 < 3) %>%
  #   pull(ID)

  # processAssays(pb[,keep], form)
  processAssays(pb, form)
  })
names(res.procList) = names(pbList)

rm(pbList, pbList2)
```


```{r dreamlet}
# dreamlet
formList = list(MSSM = update(form, ~ c02x + .),
              HBCC = update(form, ~ c08x + .))

resList = mclapply(names(res.procList), function(id){
  dreamlet(res.procList[[id]], formList[[id]])
  })
names(resList) = names(res.procList)

tab = lapply(names(res.procList), function(id){
  co = ifelse(id == "MSSM", "c02xAD", "c08xSCZ")
  tab = topTable( resList[[id]], coef=co, number=Inf)
  tab$Dataset = id
  tab %>% as_tibble
  })

df = inner_join(tab[[1]], tab[[2]], by=c("assay", 'ID'))
```

```{r plots, cache=FALSE}
df %>%
  group_by(assay) %>%
  summarize(r = cor(logFC.x, logFC.y, method="sp"),
            r.mag = cor(loess(logFC.x ~ AveExpr.x)$residuals, 
                            loess(logFC.y ~ AveExpr.y)$residuals, 
                            method="sp"))

genes = intersect(tab[[1]]$ID, tab[[2]]$ID)
ct = assayNames(pbList[[1]])

Y1 = tab[[1]] %>%
      select(assay, logFC, ID) %>%
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC") 

Y2 = tab[[2]] %>%
      select(assay, logFC, ID) %>% 
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC")

Y = inner_join(Y1, Y2, by="ID")  

C = cor(Y[,-1], use="pair")

i = grep("\\.x$", rownames(C))
j = grep("\\.y$", colnames(C))
Csub = C[i,j]
Csub = Csub[order(rownames(Csub)), order(colnames(Csub))]

breaksList = seq(-1, 1, by=.025)

pheatmap::pheatmap(Csub, 
  cluster_rows=FALSE, 
  cluster_cols=FALSE, 
  display_numbers = round(Csub,2),
  cellheight = 30, 
  cellwidth = 30,
  color=colorRampPalette(c("navy", "white", "red"))(length(breaksList)),
  breaks=breaksList,
  main = "MSSM AD/CTRL vs HBCC SCZ/CTRL (standard)")
```

```{r pca, fig.width=9}
plotPCA(res.procList[[1]], size=1.5, maxOutlierZ=15)
plotPCA(res.procList[[2]], size=1.5, maxOutlierZ=15) 
```

## MSSM
```{r outliers1}
df_out = lapply( res.procList, outlierByAssay)
names(df_out) = names(res.procList)

res = df_out[["MSSM"]] %>%
  group_by(ID) %>%
  summarize(meanZ = mean(z)) %>%
  arrange(desc(meanZ))

head(res, 10)

hist(res$meanZ)
```

## HBCC
```{r outliers2}
res = df_out[["HBCC"]] %>%
  group_by(ID) %>%
  summarize(meanZ = mean(z)) %>%
  arrange(desc(meanZ))

head(res, 10)

hist(res$meanZ)
```



# V2

```{r proc2}
# processAssays
form = ~ Sex + scale(Age) + scale(PMI) + Source + log(n_genes) + TechPC1 + TechPC2 + TechPC3 + percent_mito + mito_genes + mito_ribo + ribo_genes

# res.procList = lapply(pbList, function(pb){
#   res = processAssays(pb, ~1)

#   keep = outlierByAssay(res, nPC=5) %>%
#     group_by(ID) %>%
#     summarize(z2 = sort(z, decreasing=TRUE)[2]) %>%
#     arrange(desc(z2)) %>%
#     filter(z2 < 3) %>%
#     pull(ID)

#   processAssays(pb[,keep], form)
#   })
# names(res.procList) = names(pbList)
```

```{r dreamlet2}
# dreamlet
formList = list(MSSM = update(form, ~ c02x + .),
              HBCC = update(form, ~ c08x + .))

resList = lapply(names(res.procList), function(id){
  dreamlet(res.procList[[id]], formList[[id]])
  })
names(resList) = names(res.procList)

tab = lapply(names(res.procList), function(id){
  co = ifelse(id == "MSSM", "c02x.L", "c08x.L")
  tab = topTable( resList[[id]], coef=co, number=Inf)
  tab$Dataset = id
  tab %>% as_tibble
  })

df = inner_join(tab[[1]], tab[[2]], by=c("assay", 'ID'))
```

```{r plots2, cache=FALSE}
df %>%
  group_by(assay) %>%
  summarize(r = cor(logFC.x, logFC.y, method="sp"),
            r.mag = cor(loess(logFC.x ~ AveExpr.x)$residuals, 
                            loess(logFC.y ~ AveExpr.y)$residuals, 
                            method="sp"))
 
genes = intersect(tab[[1]]$ID, tab[[2]]$ID)
ct = assayNames(pbList[[1]])

Y1 = tab[[1]] %>%
      select(assay, logFC, ID) %>%
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC") 

Y2 = tab[[2]] %>%
      select(assay, logFC, ID) %>% 
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC")

Y = inner_join(Y1, Y2, by="ID")  

C = cor(Y[,-1], use="pair")


i = grep("\\.x$", rownames(C))
j = grep("\\.y$", colnames(C))
Csub = C[i,j]
Csub = Csub[order(rownames(Csub)), order(colnames(Csub))]

breaksList = seq(-1, 1, by=.025)

pheatmap::pheatmap(Csub, 
  cluster_rows=FALSE, 
  cluster_cols=FALSE, 
  display_numbers = round(Csub,2),
  cellheight = 30, 
  cellwidth = 30,
  color=colorRampPalette(c("navy", "white", "red"))(length(breaksList)),
  breaks=breaksList,
  main = "MSSM AD/CTRL vs HBCC SCZ/CTRL (covs)")
```

```{r exit1, cache=FALSE }
knitr::knit_exit()
```

# V3
```{r proc3}
# processAssays
form = ~ Sex + scale(Age) + scale(PMI) #+ Source + log(n_genes) + TechPC1 + TechPC2 + TechPC3 + percent_mito + mito_genes + mito_ribo + ribo_genes
res.procList = lapply(names(pbList), function(id){

  pb = pbList[[id]]

  # remove outliers
  keep = df_out[[id]] %>%
    group_by(ID) %>%
    summarize(meanZ = mean(z)) %>%
    arrange(desc(meanZ)) %>%
    filter(meanZ < 4) %>%
    pull(ID)

  processAssays(pb[,keep], form)
  })
names(res.procList) = names(pbList)
```

<!---
df_out = lapply( res.procList, outlierByAssay, nPC=5)
names(df_out) = names(res.procList)


df_out[[id]] %>%
  filter(z >5) %>%
  count(ID)


  summarize(table(ID))



keep = df_out[[id]] %>%
    group_by(assay) %>%
    mutate(fdr = p.adjust(pValue, "fdr")) %>%
    group_by(ID) %>%
    filter(fdr < 0.05) 


length(keep)



    %>%
    select(assay, ID)


    %>%
    pull(ID)

length(keep)




    summarize(meanZ = mean(z)) %>%
    arrange(desc(meanZ)) %>%
    filter(meanZ < 4) %>%
    pull(ID)

length(keep)
form = ~1
r1 = processAssays(pb, form, assays="EN")
r2 = processAssays(pb[,keep], form, assays="EN")

pdf("~/www/test.pdf")
plotPCA(r1, size=1.5, maxOutlierZ=15)
plotPCA(r2, size=1.5, maxOutlierZ=15) 
dev.off()


outlierByAssay(r1, nPC=5) %>%
  arrange(desc(z)) 


outlierByAssay(r2, nPC=2) %>%
  arrange(desc(z)) 


a = df_out[[id]] %>%
    filter(assay == "EN") %>%
    mutate(pValue = 2*pnorm(z, lower.tail=FALSE)) %>%
    pull(pValue)

--->


```{r pca2, fig.width=9}
plotPCA(res.procList[[1]], size=1.5, maxOutlierZ=15)
plotPCA(res.procList[[2]], size=1.5, maxOutlierZ=15)
```

```{r dreamlet3}
# dreamlet
formList = list(MSSM = update(form, ~ c02x + .),
              HBCC = update(form, ~ c08x + .))

resList = lapply(names(res.procList), function(id){
  dreamlet(res.procList[[id]], formList[[id]])
  })
names(resList) = names(res.procList)

tab = lapply(names(res.procList), function(id){
  co = ifelse(id == "MSSM", "c02xAD", "c08xSCZ")
  tab = topTable( resList[[id]], coef=co, number=Inf)
  tab$Dataset = id
  tab %>% as_tibble
  })

df = inner_join(tab[[1]], tab[[2]], by=c("assay", 'ID'))
```

```{r plots3, cache=FALSE}
df %>%
  group_by(assay) %>%
  summarize(r = cor(logFC.x, logFC.y, method="sp"),
            r.mag = cor(loess(logFC.x ~ AveExpr.x)$residuals, 
                            loess(logFC.y ~ AveExpr.y)$residuals, 
                            method="sp"))
 
genes = intersect(tab[[1]]$ID, tab[[2]]$ID)
ct = assayNames(pbList[[1]])

Y1 = tab[[1]] %>%
      select(assay, logFC, ID) %>%
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC") 

Y2 = tab[[2]] %>%
      select(assay, logFC, ID) %>% 
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC")

Y = inner_join(Y1, Y2, by="ID")  

C = cor(Y[,-1], use="pair")

i = grep("\\.x$", rownames(C))
j = grep("\\.y$", colnames(C))
Csub = C[i,j]
Csub = Csub[order(rownames(Csub)), order(colnames(Csub))]

breaksList = seq(-1, 1, by=.025)

pheatmap::pheatmap(Csub, 
  cluster_rows=FALSE, 
  cluster_cols=FALSE, 
  display_numbers = round(Csub,2),
  cellheight = 30, 
  cellwidth = 30,
  color=colorRampPalette(c("navy", "white", "red"))(length(breaksList)),
  breaks=breaksList,
  main = "MSSM AD/CTRL vs HBCC SCZ/CTRL (covs)")
```


# V4
```{r proc4}
# processAssays
form = ~ Sex + scale(Age) + scale(PMI) + Source + log(n_genes) + TechPC1 + TechPC2 + TechPC3 + percent_mito + mito_genes + mito_ribo + ribo_genes
res.procList = lapply(names(pbList), function(id){

  pb = pbList[[id]]

  # remove outliers
  keep = df_out[[id]] %>%
    group_by(ID) %>%
    summarize(meanZ = mean(z)) %>%
    arrange(desc(meanZ)) %>%
    filter(meanZ < 4) %>%
    pull(ID)

  processAssays(pb[,keep], form)
  })
names(res.procList) = names(pbList)
```

```{r dreamlet4}
# dreamlet
formList = list(MSSM = update(form, ~ c02x + .),
              HBCC = update(form, ~ c08x + .))

resList = lapply(names(res.procList), function(id){
  dreamlet(res.procList[[id]], formList[[id]])
  })
names(resList) = names(res.procList)

tab = lapply(names(res.procList), function(id){
  co = ifelse(id == "MSSM", "c02xAD", "c08xSCZ")
  tab = topTable( resList[[id]], coef=co, number=Inf)
  tab$Dataset = id
  tab %>% as_tibble
  })

df = inner_join(tab[[1]], tab[[2]], by=c("assay", 'ID'))
```

```{r plots4, cache=FALSE}
df %>%
  group_by(assay) %>%
  summarize(r = cor(logFC.x, logFC.y, method="sp"),
            r.mag = cor(loess(logFC.x ~ AveExpr.x)$residuals, 
                            loess(logFC.y ~ AveExpr.y)$residuals, 
                            method="sp"))
 
genes = intersect(tab[[1]]$ID, tab[[2]]$ID)
ct = assayNames(pbList[[1]])

Y1 = tab[[1]] %>%
      select(assay, logFC, ID) %>%
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC") 

Y2 = tab[[2]] %>%
      select(assay, logFC, ID) %>% 
      filter(ID %in% genes) %>%
      pivot_wider(names_from = "assay", values_from="logFC")

Y = inner_join(Y1, Y2, by="ID")  

C = cor(Y[,-1], use="pair")

i = grep("\\.x$", rownames(C))
j = grep("\\.y$", colnames(C))
Csub = C[i,j]
Csub = Csub[order(rownames(Csub)), order(colnames(Csub))]

breaksList = seq(-1, 1, by=.025)

pheatmap::pheatmap(Csub, 
  cluster_rows=FALSE, 
  cluster_cols=FALSE, 
  display_numbers = round(Csub,2),
  cellheight = 30, 
  cellwidth = 30,
  color=colorRampPalette(c("navy", "white", "red"))(length(breaksList)),
  breaks=breaksList,
  main = "MSSM AD/CTRL vs HBCC SCZ/CTRL (covs)")
```



# 1) colData() should match Samples.max
# 2) plotPCA after outliers removed

# CTRl comparson
# permuate labels
# remove TechPC's





