---
title: "Run crumblr contrasts"
subtitle: 'Automated analysis'
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.time()`"
output: 
  html_document:
    toc: true
    smart: true
    self_contained: false
params:
  dataset: NULL
  variable_type: NULL
  ctst_key: NULL
  AnnoLevel: NULL
  SampleLevel: NULL
---



<!--- 

# https://hoffmg01.hpc.mssm.edu/PsychAD_analysis/

cd /sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/analysis/freeze2/testing/
ml python git pandoc
git pull origin master

# Create jobs
Rscript write_contrast_jobs.R

# Goto folder
cd /sc/arion/projects/psychAD/NPS-AD/freeze2_rc/analysis

ls jobs/* | grep crumblr | grep HBCC | parallel -P1 "bsub < {}"

# rmarkdown::render("run_dreamlet_contrasts.Rmd")


DIR=/sc/arion/projects/CommonMind/hoffman/NPS-AD/work/nps_ad/analysis/freeze2/testing/

variable_type="CAT"
ctst_key="HBCC_SCZ_BD_CTRL"
dataset="HBCC"
OUTFILE=/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/analysis/results/${ctst_key}/${ctst_key}
Rscript -e "rmarkdown::render(\"$DIR/run_dreamlet_contrasts.Rmd\", output_file = \"$OUTFILE\", 
params=list(dataset = \"$dataset\", variable_type = \"$variable_type\", ctst_key = \"$ctst_key\"))"


params = list(dataset = 'Aging',
variable_type = 'ORD', 
SampleLevel = 'Channel', 
AnnoLevel = 'subclass', 
ctst_key = 'Aging')

--->


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning=FALSE,
  message=FALSE,
  error = FALSE,
  tidy = FALSE,
  dev = c("png", "pdf"),
  package.startup.message = FALSE,
  cache = FALSE,
  cache.lazy = FALSE)
```

# Parameters

<font size="4">__Evaluating dataset:__ `r params$dataset` </font>

<font size="4">__Variable type:__ `r params$variable_type`</font>

<font size="4">__Contrast:__ `r params$ctst_key`</font>

<font size="4">__AnnoLevel:__ `r params$AnnoLevel`</font>

<font size="4">__SampleLevel:__ `r params$SampleLevel`</font>


# Construct regression contrasts
Initialize with this line of code to reproduce results
`
```{r init, echo=FALSE, results="asis"}
res = with(params, paste0('params = list(dataset = "', dataset, '",\n', 
  'variable_type = "', variable_type, '",\n', 
  'SampleLevel = "', SampleLevel, '",\n', 
  'AnnoLevel = "', AnnoLevel, '",\n', 
  'ctst_key = "', ctst_key, '")'))
cat(res)
```
`


```{r get.contrast}
# Load CONTRASTS and metadata
load("/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/analysis/contrasts_for_dreamlet.Rdata")

suffix2 = with(params, gsub(paste0(dataset, "/", dataset, "/"), paste0(dataset, "/"), paste0(toupper(dataset), "/", ctst_key, "/", SampleLevel, "/", AnnoLevel)))

outpath = paste0("/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/analysis/results/", suffix2)

ctr = CONTRASTS[[params$variable_type]][[params$ctst_key]]

# Remove contrasts that fail
if( ctr$variable == "c07x"){
  ctr$contrasts = ctr$contrasts[!grepl("c07xBD", ctr$contrasts)]
}

# make this a random effect
ctr$covariates_incl = gsub("Brain_bank", "(1|Brain_bank)", ctr$covariates_incl)
```







# Load data
```{r load}
suppressPackageStartupMessages({
library(dreamlet)
library(SingleCellExperiment)
library(tidyverse)
library(crumblr)
library(qvalue)
library(cowplot)
library(ggcorrplot)
library(knitr)
library(aplot)
library(ggtree)
library(RColorBrewer)
library(tidyverse)})

# read processed data
path = paste0("/sc/arion/projects/psychAD/NPS-AD/freeze2_rc/pseudobulk/")
pattern = paste0(toupper(params$dataset), "_2023-06.*_PB_", params$SampleLevel, "_", params$AnnoLevel, ".RDS")
file = dir(path, pattern=pattern, full.names=TRUE)

pb = readRDS( file )

covariates_base = c("scale(Age)", "(1|Sex)", "scale(PMI)") 
if( params$SampleLevel == "Channel" ){
  covariates_base = c(covariates_base, "(1|SubID)", "(1|poolID)")
}
```

# crumblr analysis
```{r analysis, message=TRUE}
# combine base covariance with testing variables
covariates = setdiff(union(covariates_base, ctr$covariates_incl), ctr$covariates_excl)

# user-specified formula
if( any(is.null(ctr$contrasts)) || any(is.na(ctr$contrasts)) ){
  form = paste0("~ ", ctr$variable, " + ", paste0(covariates, collapse=" + "))
  form.vp = form
}else{
  form = paste0("~ 0 + ", ctr$variable, " + ", paste0(covariates, collapse=" + "))
  form.vp = paste0("~ (1|", ctr$variable, ") + ", paste0(covariates, collapse=" + "))
}
form = as.formula(form)
form.vp = as.formula(form.vp)

if( params$SampleLevel == "Channel"){
  # merge subject-level metadata with colData(res.proc)
  metadata_sub = metadata[metadata$SubID %in% colData(pb)$SubID,]
  idx = match( colData(pb)$SubID, metadata_sub$SubID)
}else{  
  metadata_sub = metadata[metadata$SubID %in% rownames(colData(pb)),]
  idx = match( rownames(colData(pb)), metadata_sub$SubID)
}

# # test ordering
# df = cbind(as.character(colData(pb)$SubID), 
#   metadata_sub$SubID[idx])
# all.equal(df[,1], df[,2])

cols = colnames(metadata_sub) %in% colnames(colData(pb))
colData(pb) = cbind(colData(pb), metadata_sub[idx,!cols])

# crumblr + dream
cobj = crumblr(cellCounts(pb))

# filter samples for this variable
idx = !is.na(colData(pb)[,ctr$variable])
```


```{r conditional_block, eval=(sum(idx) == 0)}
# if no samples retained, exit
knitr::knit_exit()
```


```{r test2}
# variance partitioning
vp.c = fitExtractVarPartModel(cobj[,idx], form.vp, colData(pb)[idx,])

fig.vp = plotPercentBars( sortCols(vp.c) )

plotVarPart(sortCols(vp.c), label.angle=60, ncol=4 ) + theme(aspect.ratio=1)

if( ! is.null(ctr$contrasts) & !any(is.na(ctr$contrasts)) ){
  L = makeContrastsDream(form, as.data.frame(colData(pb)), contrasts = ctr$contrasts )
}else{
  L = NULL
  ctr$contrasts = ctr$variable
}

fit = dream( cobj, form, colData(pb), L = L)
fit = eBayes(fit)

# get coeffs for ordinal test
if( params$variable_type == "ORD" ){
  ctr$contrasts = grep(paste0("^", ctr$contrasts, "."), colnames(coef(fit)), value=TRUE)
}
```

### Formulas

formula: 
```{r echo=FALSE} 
paste0(as.character(form), collapse=' ')
```

formula vp: 
```{r echo=FALSE} 
paste0(as.character(form.vp), collapse=' ')
```

n samples: `r ncol(cobj)`
n retained: `r nrow(fit$design)`

# Results
```{r topTable, results="asis"}
res = lapply(ctr$contrasts, function(ctst){
  topTable(fit, coef=ctst, number=Inf) %>%
      select(logFC, AveExpr,  t, P.Value, adj.P.Val) %>%
      kable(caption = ctst) 
  
})
names(res) = ctr$contrasts

if( params$variable_type == "ORD" ){
  topTable(fit, coef=ctr$contrasts, number=Inf) %>%
      select(AveExpr, F, P.Value, adj.P.Val)  %>%
      kable(caption = paste(ctr$contrasts, collapse = " + "))
}
```

# Plot of top association
```{r plot.line}
tab = topTable(fit, coef=ctr$contrasts[1], number=Inf) 
id = rownames(tab)[1]

# Plot observed CLR transform of top hit
#----------------------------------------
data = data.frame(respValue = cobj$E[id,], colData(pb)[,all.vars(form)])
data$xvalue = data[[ctr$variable[1]]]

fig1 = ggplot(data, aes(xvalue, respValue)) +
    theme_classic() +
    theme(aspect.ratio=1, legend.position="none", plot.title = element_text(hjust = 0.5)) + 
    xlab(ctr$variable[1]) + 
    ylab(paste(id, '(CLR transform)')) +
    ggtitle(ctr$contrasts[1]) 

if( params$variable_type %in% c("ORD", "CAT") ){
  fig1 = fig1 + 
    geom_boxplot(aes(fill=xvalue)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust=1))
}else{
  fig1 = fig1 + 
    geom_point()
}

# Plot CLR residuals of top hit
#------------------------------
form2 = as.formula(paste('~ 0 + ', ctr$contrasts[1]))

resids = fitVarPartModel(cobj, form, data, fxn = function(fit){
  residuals(fit) + get_prediction( fit, form2)
  })

data2 = merge(data, data.frame(resids = resids[[id]]), by="row.names")

fig2 = ggplot(data2, aes(xvalue, resids)) +
    theme_classic() +
    theme(aspect.ratio=1, legend.position="none", plot.title = element_text(hjust = 0.5)) + 
    xlab(ctr$variable[1]) + 
    ylab(paste(id, '(CLR residuals)')) +
    ggtitle(ctr$contrasts[1]) 

if( params$variable_type %in% c("ORD", "CAT") ){
  fig2 = fig2 + 
    geom_boxplot(aes(fill=xvalue)) 
}else{
  fig2 = fig2 + 
    geom_point()
}

plot_grid(fig1, fig2, nrow=1)
```






### Multivariate test along hierarchy
```{r sd, fig.width=15, fig.height=8}
hc = buildClusterTreeFromPB(pb)

figList = lapply(ctr$contrasts, function(ctst){

  res = treeTest( fit, cobj, hc, coef=ctst)

  fig1 = plotTreeTest(res) + theme(legend.position="none") + ggtitle(ctst)

  tab = topTable(fit, coef=ctst, number=Inf)

  tab$celltype = factor(rownames(tab), rev(get_taxa_name(fig1)))
  tab$se = with(tab, logFC/t)

  fig2 = ggplot(tab, aes(celltype, logFC)) + 
    geom_hline(yintercept=0, linetype="dashed", color="grey", size=1) +
    geom_errorbar(aes(ymin = logFC - 1.96*se, ymax = logFC + 1.96*se), width=0) +
    geom_point(color="dodgerblue") +
    theme_classic() +
    coord_flip() +
    xlab('') + 
    ylab("Effect size") +
    theme(axis.text.y=element_blank(),
          axis.ticks.y=element_blank())

  # combine plots
  fig.combine = fig2 %>% insert_left(fig1) %>% insert_right(fig.vp)

  # correlation 
  # get covariance matrix
  V = vcov(fit, cobj, coef=ctst)

  # convert to correlation matrix
  V = cov2cor(V)

  # shorten names
  rownames(V) = gsub(paste0(":", ctst), '', rownames(V))
  colnames(V) = gsub(paste0(":", ctst), '', colnames(V))

  # sort to match tree
  i = match(rev(get_taxa_name(fig1)), rownames(V))

  fig.cor = ggcorrplot(V[i,i], outline.color=NA, colors=c("blue3", "white", "red3")) + theme(plot.title = element_text(hjust = 0.5)) + ggtitle(ctst) 

  list(fig.combine = fig.combine, fig.cor = fig.cor)
})

lapply(figList, function(x) x$fig.combine)
lapply(figList, function(x) x$fig.cor)
```


# Save results
```{r save.results}
file = paste0(outpath, "/fit_crumblr.RDS")
saveRDS(fit, file=file)
```






Session Info
<details>
```{r session, echo=FALSE}
sessionInfo()
```
</details>


